Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO: a Bayesian study

### 2016-04-14 Paper/paper.tex
fixed typo in refs

Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}-\ref{fig:TT_LambdaMedian_vs_N_AllInOne}.

@@

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization) about the
measurability of $\lambdans$. We mitigate this by drawing multiple
realizations of population generation (i.e. multiple identically constructed
populations). Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} shows the first
results for multiple populations. Lets focus on the left panel first. We
show the median measured value of $\lambdans$ from $100$ independent 
populations with fixed $\lambdans=2000$ as the top set of curves. The one 
highlighted (in green) curve corresponds to the population discussed above.
Similarly, we generate sets of $100$ populations with $\lambdans=500-1500$, and
show the median measured $\lambdans$ as four distinct groups of curves in
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} (with the single highlighted
curve in each group corresponding to
Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}-\ref{fig:TT_LambdaMedian_vs_N_AllInOne}.
Dash-dotted horizontal lines demarcate $\pm50\%$ error intervals around
the true $\lambdans$ values. All of these populations are drawn under
paradigm A (i.e. with BH masses allowed to sample the mass-gap that spans
$2-5M_\odot$).
% 
From the figure, we can conclude with a high likelihood that our measured
$\lambdans$ values will be within $10\%$ of the real value in $15-20$ events
for less deformable neutron stars with $\lambdans\leq 1000$, and in as few as
$10$ events for more deformable stars with $\lambdans\geq 1500$.
% 
The right panel of the figure is identical to the left panel, but drawn under
population paradigm B (no BHs alloewd within the mass-gap). Under this
paradigm, we expectedly find that information accumulation is slower. It would
take $20-40$ detections with $\rho\geq10$ under this paradigm, for our (median)
measured $\lambdans$ to lie within $10\%$ of the true value (fewer for 
more deformable NSs, as one would expect).
% 
This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

Further, we investigate the statistical uncertainties associated with
$\lambdans$ measurements. We use $90\%$ confidence intervals as our measure of
the same. The left column of 
Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne} shows results
under paradigm A, where BH mass priors include the mass-gap. In each panel
we show the width of the $90\%$ confidence interval measured from $N$ events
(with $N$ on the $x$-axis), normalized by the true value of $\lambdans$ for
the population. In grey are shown results from each of $100$ population
realizations, with the single population highlighted in color corresponding
to the one we focused on in 
Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}-\ref{fig:TT_LambdaMedian_vs_N_AllInOne}.
The inset shows the same, except that the ordinate is {\it not} normalized by
the true value of $\lambdans$. From top to bottom, the population $\lambdans$
decreases from $\lambdans=2000\rightarrow 500$, corresponding to decreasingly
deformable NSs with softer equations of state. We observe the following:
(i) for moderately-hard to hard equations of state with $\lambdans\geq 1000$,
we can constrain $\lambdans$ within $\pm 50\%$ using only $10-20$ events, and
within $\pm 25\%$ with $25-40$ events; (ii) for softer equations of state with
$\lambdans<1000$, we will achieve the same accuracy with $20-30$ and $50+$ 
events, respectively; and (iii) for the first $5$ or so observations, our
measurement confidence spans the entire prior allowed range 
$\lambdans\in[0,4000]$, as shown by the plateauing towards the left edge in
each panel's inset.
% 
In the right column, each panel panel is identical to its corresponding left
one, with the difference that populations are drawn under paradigm B, which
does {\it not} allow for BH masses to fall within the mass-gap. We find that
for NSs with $\lambdans\leq 1000$, it would take $25-40$ events to
constrain $\lambdans$ within $\pm 50\%$ and $50+$ events to constrain it
within $\pm 25\%$. This is somewhat slower than paradigm A, as is to be
expected since here we preclude the lowest mass-ratios, which correspond to
signals with largest tidal signatures. For $\lambdans>1000$ we find that we
can constrain $\lambdans$ within $\pm 50\%$ with a similar number of events as
for paradigm A, but will need more ($30-40$, as compared to $25-40$) events
to further constrain it to within $\pm 25\%$ of the true value.
% 
These are some of the primary findings of this paper.

Finally, we quantitatively explore the dependence of our statistical
uncertainties for $\lambdans$ on the number of events, as well as on the true
NS deformability itself. First, we will focus on the dependence on $N$. We
assume a power-law dependence of the form
$\delta\lambdans\propto\ 1/N^\alpha$. For each of the $100$ populations 
for each of $\lambdans=500-2000$, we compute the exponent $\alpha$ as a
function of the number of observed events $N$, and show it in 
Fig.~\ref{fig:TT_PowerLawLambdaErrorVsN}. There are $100\times5=500$ curves
on the figure, with one highlighted for each value of population's $\lambdans$.
These highlighted values are only special in the sense that they correspond to
populations discussed earlier in this section (c.f.
Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0}-\ref{fig:TT_LambdaMedian_vs_N_AllInOne}).
We immediately observe two things, (i) there is a globally similar dependence
on $N$ for all populations, and (ii) information accumulates {\it faster} than
$1/\sqrt{N}$. In fact, we find that if
$\delta\lambdans\propto\frac{1}{N^\alpha}$, $\alpha$ lines in the range
$0.7_{-0.2}^{+0.2}$.
% 
Next, we focus on the dependence of $\delta\lambdans$ on $\lambdans$ of the
population itself. As suggested by Fisher-matrix studies~\cite{Lackey:2013axa},
and as for $N$, we assume the form $\delta\lambdans\propto\lambdans^\beta$.
From each set of $100$ populations with a given $\lambdans$ value, we draw one
at random, and form a set of $5$ similarly drawn populations, one for each of
$\lambdans=\{500,800,1000,1500,2000\}$. With each set, we determine $\beta$
for different number of observed events $N$. In all, we make $100$ independent
$5-$population sets and show the value of $\beta$ measured from each in 
Fig.~\ref{fig:TT_PowerLawLambdaErrorVsLambda}. We find that the assumed
relation $\delta\lambdans\propto\lambdans^\beta$ gets fairly robust for 
larger values of $N$, with $\beta$ converging to $\beta=0.5^{+0.33}_{-0.33}$.
The fact that $0<\beta<1$ implies that the relative error
$\delta\lambdans/\lambdans$ {\it decreases} with increasing $\lambdans$, while
the absolute error {\it increases}.
% 
From these results, we conclude that the measurement uncertainty for
$\lambdans$ after $N$ observations is
\begin{equation}
 \delta\lambdans\propto \dfrac{\lambdans^{0.5^{+0.33}_{-0.33}}}{N^{0.7_{-0.2}^{+0.2}}}.
\end{equation}
We also find that while these results are inferred from paradigm A populations,
paradigm B gives very similar results.

>>>>

Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.

@@

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization) about the
measurability of $\lambdans$. We mitigate this by drawing multiple
realizations of population generation (i.e. multiple identically constructed
populations). Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} shows the first
results for multiple populations. Lets focus on the left panel first. We
show the median measured value of $\lambdans$ from $100$ independent 
populations with fixed $\lambdans=2000$ as the top set of curves. The one 
highlighted (in green) curve corresponds to the population discussed above.
Similarly, we generate sets of $100$ populations with $\lambdans=500-1500$, and
show the median measured $\lambdans$ as four distinct groups of curves in
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} (with the single highlighted
curve in each group corresponding to
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}).
Dash-dotted horizontal lines demarcate $\pm50\%$ error intervals around
the true $\lambdans$ values. All of these populations are drawn under
paradigm A (i.e. with BH masses allowed to sample the mass-gap that spans
$2-5M_\odot$).
% 
From the figure, we can conclude with a high likelihood that our measured
$\lambdans$ values will be within $10\%$ of the real value in $15-20$ events
for less deformable neutron stars with $\lambdans\leq 1000$, and in as few as
$10$ events for more deformable stars with $\lambdans\geq 1500$.
% 
The right panel of the figure is identical to the left panel, but drawn under
population paradigm B (no BHs alloewd within the mass-gap). Under this
paradigm, we expectedly find that information accumulation is slower. It would
take $20-40$ detections with $\rho\geq10$ under this paradigm, for our (median)
measured $\lambdans$ to lie within $10\%$ of the true value (fewer for 
more deformable NSs, as one would expect).
% 
This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

Further, we investigate the statistical uncertainties associated with
$\lambdans$ measurements. We use $90\%$ confidence intervals as our measure of
the same. The left column of 
Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne} shows results
under paradigm A, where BH mass priors include the mass-gap. In each panel
we show the width of the $90\%$ confidence interval measured from $N$ events
(with $N$ on the $x$-axis), normalized by the true value of $\lambdans$ for
the population. In grey are shown results from each of $100$ population
realizations, with the single population highlighted in color corresponding
to the one we focused on in 
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.
The inset shows the same, except that the ordinate is {\it not} normalized by
the true value of $\lambdans$. From top to bottom, the population $\lambdans$
decreases from $\lambdans=2000\rightarrow 500$, corresponding to decreasingly
deformable NSs with softer equations of state. We observe the following:
(i) for moderately-hard to hard equations of state with $\lambdans\geq 1000$,
we can constrain $\lambdans$ within $\pm 50\%$ using only $10-20$ events, and
within $\pm 25\%$ with $25-40$ events; (ii) for softer equations of state with
$\lambdans<1000$, we will achieve the same accuracy with $20-30$ and $50+$ 
events, respectively; and (iii) for the first $5$ or so observations, our
measurement confidence spans the entire prior allowed range 
$\lambdans\in[0,4000]$, as shown by the plateauing towards the left edge in
each panel's inset.
% 
In the right column, each panel panel is identical to its corresponding left
one, with the difference that populations are drawn under paradigm B, which
does {\it not} allow for BH masses to fall within the mass-gap. We find that
for NSs with $\lambdans\leq 1000$, it would take $25-40$ events to
constrain $\lambdans$ within $\pm 50\%$ and $50+$ events to constrain it
within $\pm 25\%$. This is somewhat slower than paradigm A, as is to be
expected since here we preclude the lowest mass-ratios, which correspond to
signals with largest tidal signatures. For $\lambdans>1000$ we find that we
can constrain $\lambdans$ within $\pm 50\%$ with a similar number of events as
for paradigm A, but will need more ($30-40$, as compared to $25-40$) events
to further constrain it to within $\pm 25\%$ of the true value.
% 
These are some of the primary findings of this paper.

Finally, we quantitatively explore the dependence of our statistical
uncertainties for $\lambdans$ on the number of events, as well as on the true
NS deformability itself. First, we will focus on the dependence on $N$. We
assume a power-law dependence of the form
$\delta\lambdans\propto\ 1/N^\alpha$. For each of the $100$ populations 
for each of $\lambdans=500-2000$, we compute the exponent $\alpha$ as a
function of the number of observed events $N$, and show it in 
Fig.~\ref{fig:TT_PowerLawLambdaErrorVsN}. There are $100\times5=500$ curves
on the figure, with one highlighted for each value of population's $\lambdans$.
These highlighted values are only special in the sense that they correspond to
populations discussed earlier in this section (c.f.
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}).
We immediately observe two things, (i) there is a globally similar dependence
on $N$ for all populations, and (ii) information accumulates {\it faster} than
$1/\sqrt{N}$. In fact, we find that if
$\delta\lambdans\propto\frac{1}{N^\alpha}$, $\alpha$ lines in the range
$0.7_{-0.2}^{+0.2}$.
% 
Next, we focus on the dependence of $\delta\lambdans$ on $\lambdans$ of the
population itself. As suggested by Fisher-matrix studies~\cite{Lackey:2013axa},
and as for $N$, we assume the form $\delta\lambdans\propto\lambdans^\beta$.
From each set of $100$ populations with a given $\lambdans$ value, we draw one
at random, and form a set of $5$ similarly drawn populations, one for each of
$\lambdans=\{500,800,1000,1500,2000\}$. With each set, we determine $\beta$
for different number of observed events $N$. In all, we make $100$ independent
$5-$population sets and show the value of $\beta$ measured from each in 
Fig.~\ref{fig:TT_PowerLawLambdaErrorVsLambda}. We find that the assumed
relation $\delta\lambdans\propto\lambdans^\beta$ gets fairly robust for 
larger values of $N$, with $\beta$ converging to $\beta=0.5^{+0.33}_{-0.33}$.
The fact that $0<\beta<1$ implies that the relative error
$\delta\lambdans/\lambdans$ {\it decreases} with increasing $\lambdans$, while
the absolute error {\it increases}.
% 
From these results, we conclude that the measurement uncertainty for
$\lambdans$ after $N$ observations is
\begin{equation}
 \delta\lambdans\propto \dfrac{\lambdans^{0.5^{+0.33}_{-0.33}}}{N^{0.7_{-0.2}^{+0.2}}}.
\end{equation}
We also find that while these results are inferred from paradigm A populations,
paradigm B gives very similar results.
